Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-28 Manuscript/manuscript.Rmd
edits to discuss and legend of figure 5

![**preciseTAD boundaries are enriched in CTCF sites.** (A) Visualizing locations of predicted vs called ARROWHEAD TAD boundaries on GM12878 (chr 21: 15150000-17270000). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox. The blue vertical lines represent boundaries identified by ARROWHEAD, with the corresponding called TADs outlined in blue on the upper diagonal contact matrix above. The black line represents the predicted probability of each base pair being a TAD boundary when validated by our random forest. The green vertical lines represent the *preciseTAD boundary points* (PTBP) predicted by *preciseTAD*, with the corresponding demarcated TADs outlined above in green. Below is a zoomed in portion of the genome showing a cluster of base pairs with high probability representing a single *preciseTAD boundary region (PTBR) colored in yellow. (B) Enriched heatmaps comparing the enrichment of CTCF sites around 5 kb flanked preciseTAD and ARROWHEAD TAD boundaries respectively. (C) Enriched heatmaps comparing the enrichment of CTCF sites around 10 kb flanked preciseTAD and PEAKACHU loop boundaries respectively. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF site between preciseTAD boundaries and both ARROWHEAD and PEAKACHU boundaries, respectively.](Figures/figure5_enrichment_and_log2dist_ctcf.png)

@@

Here we present *preciseTAD*, a random forest-based approach coupled with spatial clustering (DBSCAN) and scalable partition around medoids (CLARA) that allows for more precise prediction of TAD or loop boundaries. *preciseTAD* leverages a random forest (RF) classification algorithm built on ground truth TAD(loop) boundaries obtained from a user defined TAD(loop) calling tool and ChIP-seq defined genomic annotations as the feature space. We optimized our RF model by creating a computational framework that systematically compared the performances of different combinations of genome binning (resolution), feature engineering procedure, and re-sampling technique. In doing so, we first introduced a new strategy for establishing boundary prediction data referred to as *shifted binning*, in which the rows of the data matrix were controlled by the data resolution (either 5kb, 10kb, 25kb, 50kb, and 100kb) used to obtain ground truth boundaries, while assuring ground truth bounderies were located central to any given genomic bin. Furthermore, we introduced a new procedure for drawing spatial associations between boundary regions and genomic annotations, referred to as *distance*-type predictors. We benchmarked *preciseTAD* against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and PEAKACHU [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. For a more in depth study, we also applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from the GM12878 and K562 cell lines.

@@

Besides balanced accuracy (BA), we investigated five other performance metrics to evaluate the optimal combination of data level characteristics for boundary region prediction including accuracy, area under the receiver operating characteristic curve (AUC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Supplementary Tables 6-7). The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benifits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling.

The performance of *preciseTAD* was found to be empirically better than both ARROWHEAD and PEAKACHU ground truth boundaries. *preciseTAD* provides bp resolution predicted boundaries, thereby alleviating resolution limitations of conventional TAD-callers. We show that, unlike ARROWHEAD, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries. *preciseTAD* boundaries we found to be more enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143. Likewise *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, further highlighting their biological significance.

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined TAD(loop)-caller. Ideally we would perform simulation studies to benchmark *preciseTAD* against simulated boundaries. While there are existing methods for simulating Hi-C data sets [@forcato2017comparison; @zheng2020freehi], of which boundaries can be extracted from, to date, there are no such methods for simulating ChIP-seq data. However, given *preciseTAD* is applied on chromosome-specific sections of the linear genome, we feel that aggregating results across chromosomes provides strong empirical evidence of the superiority of *preciseTAD*. Another limitation is that *preciseTAD* does not provide hierarchical boundaries, but instead simply provides the bp coordinate that demarcates TADs(loops). Studies have shown evidence of the hierarchical nature to the 3D architecture of the human genome [@hansen2018recent; @weinreb2016identification; @fraser2015hierarchical]. In such a hierarchy, the boundaries provided by *preciseTAD* can be interpreted as the lowest level of boundaries.

In summary, we have shown that TAD boundary prediction is a multi-faceted problem composed of many data irregularities and characteristics. Simply quering a contact matrix ignores the fundamental roles of known molecular drivers of 3D chromatin. Instead, we propose a method for turning TAD-calling into a supervised learning framework built using ChIP-seq annotations. Specifically, we introduce two novel concepts known as *shifted binning* and *distance*-type predictors which we use to build random forest classification models for predicting boundary regions. We then introduce *preciseTAD*, a method for leveraging random forests to predict TAD(loop) boundaries at base pair resolution. Our method is able to bridge the resolution gap between 1D ChIP-seq annotations and 3D Hi-C sequencing data for more precise and biologically meaningful boundary prediction. By releasing our method as an open source R package, we hope to provide an efficient and easy-to-use tool to further explore the 3D organization of the human genome.

>>>>

![**preciseTAD boundaries are enriched in CTCF sites.** (A) Visualizing locations of predicted vs called ARROWHEAD TAD boundaries on GM12878 (chr 21: 15150000-17270000). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox. The blue vertical lines represent boundaries identified by ARROWHEAD, with the corresponding called TADs outlined in blue on the upper diagonal contact matrix above. The black line represents the predicted probability of each base pair being a TAD boundary when validated by our random forest. The green vertical lines represent the *preciseTAD boundary points* (PTBP) predicted by *preciseTAD*, with the corresponding demarcated TADs outlined above in green. Below is a zoomed in portion of the genome showing a cluster of base pairs with high probability representing a single *preciseTAD boundary region (PTBR) colored in yellow. (B) Enriched heatmaps comparing the enrichment of CTCF sites around 5 kb flanked preciseTAD and ARROWHEAD TAD boundaries respectively. (C) Enriched heatmaps comparing the enrichment of CTCF sites around 10 kb flanked preciseTAD and PEAKACHU loop boundaries respectively. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF site between preciseTAD boundaries and both ARROWHEAD and PEAKACHU boundaries, respectively. (E) Matrix of pairwise Jaccard indices across resolutions shows more invariance toward resolution for *preciseTAD* (upper triangular) compared to ARROWHEAD (lower triangular).](Figures/figure5_enrichment_and_log2dist_ctcf.png)

@@

Here we present *preciseTAD*, a random forest-based approach coupled with spatial clustering (DBSCAN) and scalable partition around medoids (CLARA) that allows for more precise prediction of TAD or loop boundaries. *preciseTAD* leverages a random forest (RF) classification algorithm built on ground truth TAD(loop) boundaries obtained from a TAD(loop) calling tool and ChIP-seq defined genomic annotations as the feature space. We first optimized our RF model by creating a computational framework that systematically compared the performances of different combinations of genome binning (resolution), feature engineering procedure, and re-sampling technique. In doing so, we first introduced a new strategy for establishing boundary prediction data referred to as *shifted binning*, in which the rows of the data matrix were controlled by the data resolution (either 5kb, 10kb, 25kb, 50kb, and 100kb) used to obtain ground truth boundaries, while assuring ground truth bounderies were located central to any given genomic bin. Furthermore, we introduced a new procedure for drawing spatial associations between boundary regions and genomic annotations, referred to as *distance*-type predictors, defined as as the $log_{2}$ distance in base pairs from each genomic bin the nearest ChIP-seq peak region. We benchmarked *preciseTAD* against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and PEAKACHU [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. For a more in depth study, we also applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from the GM12878 and K562 cell lines.

@@

Besides balanced accuracy (BA), we investigated five other performance metrics to evaluate the optimal combination of data level characteristics for RF prediction including accuracy, area under the receiver operating characteristic curve (AUC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Supplementary Tables 6-7). The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benifits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling.

The performance of *preciseTAD* was found to be empirically better than both ARROWHEAD and PEAKACHU ground truth boundaries. *preciseTAD* provides bp resolution predicted boundaries, thereby alleviating resolution limitations of conventional TAD-callers. We show that, unlike ARROWHEAD, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries. *preciseTAD* boundaries were found to be more enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143. Likewise *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, further highlighting their biological significance.

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined TAD(loop)-caller. Ideally we would perform simulation studies to benchmark *preciseTAD* against simulated boundaries. While there are existing methods for simulating Hi-C data sets [@forcato2017comparison; @zheng2020freehi], of which boundaries can be extracted from, to date, there are no such methods for simulating ChIP-seq data. However, given *preciseTAD* is applied on chromosome-specific sections of the linear genome, and on multiple cell lines, we feel that aggregating results across chromosomes provides strong empirical evidence of the superiority of *preciseTAD*. Another limitation is that *preciseTAD* does not provide hierarchical boundaries, but instead simply provides the bp coordinate that demarcates TADs(loops). Studies have shown evidence of the hierarchical nature to the 3D architecture of the human genome [@hansen2018recent; @weinreb2016identification; @fraser2015hierarchical]. In such a hierarchy, the boundaries provided by *preciseTAD* can be interpreted as the lowest level of boundaries.

In summary, we have shown that TAD boundary prediction is a multi-faceted problem composed of many data irregularities and characteristics. Simply quering a contact matrix ignores the fundamental roles of known molecular drivers of 3D chromatin. Instead, we propose a method for turning TAD-calling into a supervised learning framework built using ChIP-seq annotations. Specifically, we introduce two novel concepts known as *shifted binning* and *distance*-type predictors which we use to build random forest classification models for predicting boundary regions. We then introduce *preciseTAD*, a method for leveraging random forests to predict TAD(loop) boundaries at base pair resolution. Our method is able to bridge the resolution gap between 1D ChIP-seq annotations and 3D Hi-C sequencing data for more precise and biologically meaningful boundary identification. By releasing our method as an open source R package, we hope to provide an efficient and easy-to-use tool to further explore the 3D organization of the human genome.
